---
title: "Constructing and fitting admixture graphs to D statistics"
author: "Thomas Mailund"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constructing and fitting admixture graphs to D statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r preamble, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admixturegraph)
```

This document describes the `admixturegraph` package, a package for fitting admixture graphs to genetic data.

The package enables you to specify admixture graphs, extract the implied equations a given graph makes for predicting correlations of genetic drift as specified by the expected $f_2$, $f_3$ and $f_4$ statistics, or fit a graph to data summarized as $D$ statistics and test the goodness of fit.

The package does not (yet) support graph inference or functionality for exploring the graph space over a set of populations.



# Admixture graphs and genetic drift

Gene frequencies in related populations are correlated. If two populations split apart a certain time in the past, all gene frequencies were the same at that split time---where the populations were one and the same---and has since drifted apart in the time since---where the populations have evolved independently.

With three related populations where one branched off a common ancestor early and the other two later in time the two closest related populations will, all else being equal, have more correlated gene frequencies than they have with the third population. 

All else being equal here means that the populations should have evolved at roughly the same speed---the gene frequency changes should not be more dramatic in one population than another. If we consider populations A, B and C, with A and B closest relatives but where population A has changed very rapidly since it split from population B, then gene frequencies in A and B could be less correlated than between B and C (but A and C would be even less correlated).

```{r, echo=FALSE}
leaves <- c("A", "B", "C")
inner_nodes <- c("AB", "ABC")
edges <- parent_edges(c(edge("A", "AB"),
                        edge("B", "AB"),
                        edge("AB", "ABC"),
                        edge("C", "ABC")))
graph <- agraph(leaves, inner_nodes, edges, NULL)
plot(graph)  
```

The way gene frequencies change over time as a population evolve is called genetic drift and the way gene frequencies are correlated between populations is determined by how they are related. These relationships can be simple trees with ancestral populations split into descendant populations or be more complex with gene flow between diverged populations.

Assuming that populations evolve by either splitting into descendant populations or merging in admixture events---but do not experience periods of continuous gene flow---their relationship can be described as a so-called *admixture graph* [@Patterson:ADMIXTOOLS]. Such a graph not only describes the history of the populations but can also be seen as a specification of how gene frequencies in the populations will be correlated. Different summary statistics can capture the correlations in gene frequencies between populations, extracting different aspects of this correlation, and the expected values of these statistics can be computed as polynomials over branch lengths and admixture proportions in the admixture graph.


# Building admixture graphs

In the `admixturegraph` package, graphs are build by specifying the edges in the graph and naming the admixture proportions so these have a variable associated.

It is first necessary to specify the nodes in the graphs, where there is a distinction between leaves and inner nodes so leaves later can correspond to values in $f$ statistics computed from genetic data. It is necessary to name the nodes so we have something to refer to when specifying the edges.

Edges are specified by specifying the parent of each node. Only one node is allowed not to have a parent; the package cannot deal with more than one root.

The tree shown above is specified as:

```{r}
leaves <- c("A", "B", "C")
inner_nodes <- c("AB", "ABC")
edges <- parent_edges(c(edge("A", "AB"),
                        edge("B", "AB"),
                        edge("AB", "ABC"),
                        edge("C", "ABC")))
graph <- agraph(leaves, inner_nodes, edges, NULL)
```

and plotted like:

```{r}
plot(graph)  
```

By default the labels on inner nodes are not shown but this can be changed by:

```{r}
plot(graph, show_inner_node_labels = TRUE)  
```

The left-to-right order of leaves in plots is always the same as the order given when constructing the graph; there is currently no good layout algorithm implemented for plotting so this is the only way of controlling the plots.

Admixture events are specified by having admixture edges. These are edges with two parents. Admixture edges really represent two different edges, capturing the drift between the two populations ancestral to the admixture event before the admixture took place, so we typically have to add inner nodes on the edges where these two ancestral populations branched off other lineages in the graph.

```{r}
leaves <- c("A", "B", "C")
inner_nodes <- c("a", "c", "ABC")
edges <- parent_edges(c(edge("A", "a"), edge("a", "ABC"),
                        edge("C", "c"), edge("c", "ABC"),
                        admixture_edge("B", "a", "c")))
graph <- agraph(leaves, inner_nodes, edges, NULL)

plot(graph, show_inner_node_labels = TRUE)
```

Since we usually model admixtures that happened at some point in the past we usually also need an inner node above the present day admixed population.

```{r}
leaves <- c("A", "B", "C")
inner_nodes <- c("a", "c", "b", "ABC")
edges <- parent_edges(c(edge("A", "a"), edge("a", "ABC"),
                        edge("C", "c"), edge("c", "ABC"),
                        edge("B", "b"),
                        admixture_edge("b", "a", "c")))
graph <- agraph(leaves, inner_nodes, edges, NULL)

plot(graph, show_inner_node_labels = TRUE)
```

This graph doesn't contain a variable for the admixture proportions between the two ancestral populations to B. To work with the equations for expected drift statistics we need to specify this as well.

This is done through the last parameter when constructing the admixture graph, like this:

```{r}
leaves <- c("A", "B", "C")
inner_nodes <- c("a", "c", "b", "ABC")
edges <- parent_edges(c(edge("A", "a"), edge("a", "ABC"),
                        edge("C", "c"), edge("c", "ABC"),
                        edge("B", "b"),
                        admixture_edge("b", "a", "c")))
admixtures <- admixture_proportions(c(
    admix_props("b", "a", "c", "alpha")
    ))
graph <- agraph(leaves, inner_nodes, edges, admixtures)

plot(graph, show_inner_node_labels = TRUE, show_admixture_labels = TRUE)
```

## A more complex example



# Extracting drift equations for expected $f$ statistics

# Fitting admixture graphs to D statistics

# References
